---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
# Reproducible Research Week 2 Assignment 1 by Metehan Soysal
This is the document that explains the steps taken to prepare the assingment for Reproducible Research Asssigment 1.

## Loading and preprocessing the data
### Libraries and Loading the Data
In order to complete the assignment we need to first load the required libraries and the data.

```{r}
library(lubridate)
library(ggplot2)
library(dplyr)
act_data <- read.csv("activity.csv", header = TRUE, sep = ',', colClasses = c("numeric", "character", "integer"))
```

### Tidying the Data
We need to convert the date column from character to date format. We will use lubridate package and ymd function for that.
```{r}
act_data$date <- ymd(act_data$date)
```

## What is mean total number of steps taken per day?
### Calculate the steps taken by day
We will first calculate the steps taken per day and will use group_by function to do that. We also need to omit NA values.

```{r}
daily_steps <- act_data %>%
  na.omit() %>%
  group_by(date) %>%
  summarize(daily_steps = sum(steps))
  print(daily_steps)
```
### Constructing the Plot
We will use ggplot library to create the plot.
```{r}
daily_plot<- ggplot(daily_steps, aes(x=daily_steps)) + geom_histogram(fill = "firebrick", binwidth = 1000) + xlab("Todal Number of Steps per Day") + ylab("Frequency")
print(daily_plot)
```

### Calculating the Mean and the Median
```{r}
mean_daily_steps <- mean(daily_steps$daily_steps, na.rm = TRUE)
mean_daily_steps

median_daily_steps <- median(daily_steps$daily_steps, na.rm = TRUE)
median_daily_steps
```

## What is the average daily activity pattern?
### Calculate the Average Steps
```{r}
interval_steps <- act_data %>% 
  na.omit() %>% 
  group_by(interval) %>% 
  summarize(steps = mean(steps))
head(interval_steps)
```
### Plotting the Average Daily Activity
```{r}
avg_steps <- ggplot(interval_steps, aes(x=interval, y=steps)) + geom_line(color = "firebrick")
print(avg_steps)
```

### Finding the Maximum Steps Occurrence in a Time Interval
```{r}
interval_steps[which.max(interval_steps$steps),]
```

## Imputing missing values
### Calculating the Missing Values
```{r}
sum(is.na(act_data$steps))
sum(is.na(act_data$interval))
sum(is.na(act_data$date))
```

### Fill-in the missing values
We will use the interval average steps to fill in for the missing values.
```{r}
act_data_filled <- cbind(act_data, interval_steps$steps)
names(act_data_filled) <- c("steps", "date", "interval", "average")

for (i in 1:length(act_data_filled$interval)) {
  if (is.na(act_data_filled[i,1])){
    act_data_filled[i,1] <- act_data_filled[i,4]}
  }
```

### Re-drawing the histogram with the filled values
```{r}
daily_steps_filled <- act_data_filled %>%
      na.omit() %>%
      group_by(date) %>%
      summarize(daily_steps = sum(steps))
print(daily_steps_filled)

daily_plot_filled <- ggplot(daily_steps_filled, aes(x=daily_steps)) + geom_histogram(fill = "firebrick", binwidth = 1000) + xlab("Todal Number of Steps per Day") + ylab("Frequency")
print(daily_plot_filled)
```

Calculating the mean and the median of the full data set.
```{r}
mean_daily_steps_filled <- mean(daily_steps_filled$daily_steps)
mean_daily_steps_filled

median_daily_steps_filled <- median(daily_steps_filled$daily_steps)
median_daily_steps_filled
```


## Are there differences in activity patterns between weekdays and weekends?
We will create extra variables to find out if the date represents a weekday or a weekend.
First we calculate the actual day of the week which date represents.
```{r}
act_data_filled$day <- weekdays(act_data_filled$date)
```
We will then use this function below to find out whether this is a weekend or a weekday.
```{r}
daytype <- function(day) {
  if (day %in% c("Saturday", "Sunday")){"weekend"} 
  else {"weekday"}
  }
```
We will then create a new column for each date to find out whether it is a weekday or a weekend.
```{r}
act_data_filled$daytype <- as.factor(sapply(act_data_filled$day, daytype))
table(act_data_filled$daytype)
```

Now we will calculate the average steps in the intervals
```{r}
interval_filled <- act_data_filled %>%
  group_by(interval, daytype) %>%
  summarise(steps = mean(steps))
```

We will go ahead and create a plot to show the differences between weekdays and weekends.
```{r}
interval_filled_plot <- ggplot(interval_filled, aes(x=interval, y=steps, color = daytype)) + geom_line() + facet_wrap(~daytype, ncol =1, nrow=2)
print(interval_filled_plot)
```


